%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}

  First and foremost, I would like to thank my lead supervisor, Prof.~\'{O}scar, for all the guidance, irrespective of day
time and circumstances, for all the unconditional friendship, and for the constant flood of sheer knowledge that was
so valuable throughout this process.
To Prof.~Filipe Silva, for giving me the chance of working on this
incredible project, supported by Portuguese Foundation of Science and Technology
national funds, under the national support to R\&D units grant, through the
reference projects UIDB/04436/2020 and UIDP/04436/2020, and also, through the
project `HAMaBICo --- Hybrid Additive Manufacturing for Bio-Inspired
Components', reference NORTE-01-0145-FEDER-000018 and by the project
Add.Additive\_Manufacturing to Portuguese Industry\_POCI-01-0247-FEDER-024533.

To Paulo Pinto for not only being a stupendous colleague, but a true
friend, hearing my constant complaints and providing the assertive advice. To
Flávio for the manufacturing tests and for the valuable inputs given. To
Joseph Lennox (@lordofthehypens) and the open-source project \texttt{Slic3r} for the incredible help in the slicer and path generator development.

To my parents for always pushing me forward and for being the life example of character and sheer determination, a
lamp lit on a stormy day. Thank you also for an upbringing full of debates and
argumentation, which has undoubtedly contributed for mastering the challenges I
took ahead,
%and for instilling in me the taste in good music which helped me
%throughout this tortuous journey, listening to the \emph{Songs of Distant
%  Earth},
and sorry for my constant lack of time.
To my sister, for always being so supportive and caring, helping out with your
immense ingenuity and creativity, a mix between the engineer and the artist.

To my closest friends, Pedro, Guinho, Machado e Raúl for taking me out to dinner
and drink, which is exactly what one needs sometimes. For all the support and
the true friendship that survives even the end of times. To my beloved dog
Pantufa for adopting me eight years ago when I was going crazy with my
first MSc thesis and never leaving my side.

Lastly, it is important to acknowledge that the present work extends a previous
one, done in parallel, yielding the first version of a multi-material LPBF
equipment~\cite{piresMecat}. However, the present work addresses all the
ecosystem, and not only the equipment.

%Acknowledgments are personal text and should be a free expression of the author.
%
%However, without any intention of conditioning the form or content of this text, I would like to add that it usually starts with academic thanks (instructors, etc.); then institutional thanks (Research Center, Department, Faculty, University, FCT / MEC scholarships, etc.) and, finally, the personal ones (friends, family, etc.).
%
%But I insist that there are no fixed rules for this text, and it must, above all, express what the author feels.

\end{ntacknowledgements}