%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}

First and foremost, I would like to thank my supervisor, Prof.~Sandro Pinto, for
all the guidance, support, and technical expertise that was so valuable
throughout this process.

I also would like to thank my lab colleagues Martins and David for all the help
with the Bao hypervisor and for all the patience and wise advice, without which
this work would not have been possible.

To my parents for always pushing me forward and for being the life example of character and sheer determination, a
lamp lit on a stormy day. Thank you also for an upbringing full of debates and
argumentation, which has undoubtedly contributed to mastering the challenges I
took ahead, and for instilling in me the taste in good music which helped me
throughout this journey, listening to the \emph{Wish You Were Here}. Thank you
for all the support, love, friendship, and for always believing in me,
and sorry for my constant lack of time.
To my sister, for always being so supportive and caring, helping out with your
immense ingenuity and creativity, a mix between the engineer and the artist.

To my closest friends, Pedro and Chicão for taking me out to dinner, which is
exactly what one needs sometimes.
For all the support and the true friendship that survives even the end of times.
To Pedro for giving me the opportunity of
being a godfather to a wonderful girl, Matilde. I hope I made you proud
too. To my beloved dog
Pantufa for adopting me eleven years ago when I was going crazy with my
first MSc thesis and never leaving my side.

And last, but most importantly, to my beautiful and amazing girlfriend, Rafaela,
my one true love. We've faced so many challenges, the distance and the lack of
time, but we strived, and, like a boat that emerges out of a storm, we came
stronger and more vibrant, full of life. And now we can appreciate the
petrichor from the summer rains together. Your unconditional love and support meant the world
to me, from day one to today, and through the many more we'll face
together. I hope you are proud of me, because this accomplishment is also
yours. Love you so damn much!!!

\end{ntacknowledgements}
